### 2016-07-05 image_setup.tex
Finishing image local mount edits

\begin{lstlisting}[]
$ (echo d; echo ; echo n; echo p; echo ; echo 137216; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The \texttt{echo} commands within the parenthesis are interpreted as
key-presses in the \texttt{fdisk} commmand mode. They (i) delete the last
partition, (ii) create a new primary partition, (iii) print new partition
table and (iv) write new partition table to the image file. Use a loopback
device to make the Raspbian image available as a block device:

@@

\texttt{lsblk} may be used to list block devices. It appears that the image is
available as \texttt{/dev/loop0} which has two partitions \texttt{loop0p1} and
\texttt{loop0p2}:
%Use \texttt{lsblk} to view the partitions:

@@

Check and resize the root partition:
\begin{lstlisting}[]
# e2fsck -f ${DEV}p2
# resize2fs ${DEV}p2
\end{lstlisting}
\FloatBarrier

%We see that the boot and root partitions starts at sector 8192 and 137216 respectivly.
%We also notice that each sector is 512 bytes.
%On newer systems, we can mount the device easier using "losetup --show -f -P IMAGE"
%Mount root partition:

\subsection{Mount image}

Browsing and altering the files in the image is easy. Simply mount the partitions.
%To browser and alter the files in the image, it is possible to mount the partitions.

Lets first mount the root partition. This is done by creating an empty directory
that we will be used as a mountpoint. We name it \texttt{root} and create it in the
working directory before mounting the root filesystem onto the mountpoint.
%After this, mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

@@

The root filesystem already has a boot directory that can be used as
the mountpoint for the boot partition.
%After this, we can also mount the boot partition inside the newly mounted
%root filesystem.
This is convenient because that partition will be mounted
on this directory when a \ac{Raspi} starts up with a memory card containing
the image. %It is therefore the natural place to mount it for later purposes.
Now, mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

%We can now change all the files in the disk image as desired.
It is now possible to change all files within the Raspbian image as desired.
We will take advantage of this to edit configuration files, append new files,
and even update and install packages.

>>>>

\begin{lstlisting}[]
$ (echo d; echo 2; echo n; echo ; echo ; echo 137216; echo ; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The \texttt{echo} commands within the parenthesis are interpreted as
key-presses in the \texttt{fdisk} commmand mode. They (i) delete partition
number 2, (ii) create a new primary type partition by default, (iii) set the
new partition starting point (137216) and (iv) write new partition table to
the image file. The previously shown command options of \texttt{fdisk}
should work for both old and new versions of the commands. However, if this
is not the case, the testbed administrator can do these commands manually.
In this case, it is only required to delete the old root
filesystem partition and create the new one starting from the same sector
number. In our case, if the partitions commands were correct, the output
should be following:

\begin{lstlisting}[]

Command (m for help): Partition number (1-4):
Command (m for help): Partition type:
   p   primary (1 primary, 0 extended, 3 free)
   e   extended
Select (default p): Using default response p
Partition number (1-4, default 2): Using default value 2
First sector (2048-4806655, default 2048): Last sector, +sectors or
+size{K,M,G} (137216-4806655, default 4806655): Using default value 4806655

Command (m for help): The partition table has been altered!

Syncing disks.

\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\subsection{Loopback Device Setup}
After successfully resizing the image file, we use a loopback device to make
the Raspbian image available as a block device in the filesystem. For this
command to work, the tested administrator distribution must have the
\texttt{utils-linux} package with version 2.21 or higher. Otherwise, the
\texttt{-P} argument of \texttt{losetup} will appear as invalid.

@@

If the previous command was succesful, the \texttt{lsblk} command can be used
to list the available block devices in the filesystem as follows:
%Use \texttt{lsblk} to view the partitions:

@@

The image block device appears as \texttt{/dev/loop0}. This block device has
two partitions associated to it, e.g. \texttt{loop0p1} and \texttt{loop0p2}.
Finally, we check the block device status with \texttt{e2fsck} and
\texttt{resize2fs}, respectively in the command block:

\begin{lstlisting}[]
# e2fsck -f ${DEV}p2
e2fsck 1.42.8 (20-Jun-2013)
Pass 1: Checking inodes, blocks, and sizes
Pass 2: Checking directory structure
Pass 3: Checking directory connectivity
Pass 4: Checking reference counts
Pass 5: Checking group summary information
/dev/loop0p2: 35392/80480 files (0.1% non-contiguous), 201968/321536 blocks
# resize2fs ${DEV}p2
resize2fs 1.42.8 (20-Jun-2013)
Resizing the filesystem on /dev/loop0 to 583680 (4k) blocks.
The filesystem on /dev/loop0 is now 583680 blocks long.
\end{lstlisting}
\FloatBarrier

\subsection{Block Devices Mounting}
For browsing and altering the files in the image, we mount the block
device partitions into a particular path of our \texttt{WORKDIR} in order
to customize them. We proceed first to mount the block device partition that
contains the root filesystem and later the boot partition. This is done by
creating an empty directory that it is used as a mountpoint. We name it
\texttt{root} and create it in the working directory before mounting the
root filesystem onto the mountpoint. Thus, we mount the root filesystem
as follows:

@@

The root filesystem mounted in \texttt{ROOTDIR}, already has a boot
directory that can be used as the mount point for the boot partition
in the block device \texttt{loop0p1}. This is convenient because
the final edited partition from \texttt{ROOTDIR/boot} will be mounted on
this same directory when a \ac{Raspi} starts up with a memory card
containing the raw final image. Hence, to mount boot partition we do:
%It is therefore the natural place to mount it for later purposes.

\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot
%We can now change all the files in the disk image as desired.

In this way, it is now possible to change all files within the Raspbian
image as desired by editing the files in \texttt{ROOTDIR}. We take
advantage of this to edit configuration files, append new files and even
update and install packages.

\subsection{Configuring OS files and scripts}

The \ac{Raspi}s should be setup as similar as possible. One of those
few settings that may be different in the devices is their hostname.
%The hostname helps the user to distinguish the devices from each other.
%The hostname helps distinguish them from each other.
One approach to create unique hostnames would be to write the hostname
in each memory card. Another way is to create a start up script to automatically
configure the hostname. We decided to define the hostnames based on the \ac{MAC}
addresses of the \ac{Raspi}s wired ethernet interface. We store the \ac{MAC}
addresses and hostnames of the \ac{Raspi}s in a file as below. The testbed
administrator will need to insert the \ac{MAC} addresses and hostnames of
his/her \ac{Raspi}s.

%The \ac{Raspi}s should be setup to be as similar as possible. One of those
%things that is preferred to be different on the devices is their hostname.
%This helps distinguish them from each other. One solution to this in our
%approach would be to uniquely change the hostname in each memory card.
%This is however not desired because that would either require modifying
%the customized Raspbian lite image before writing the content to the
%memory card or to manually change it in the \ac{Raspi}s after the image has
%been written to a memory card.
%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.
%
%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames. This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

%Instead, we decided to register the ethernet MAC addresses of all the \ac{Raspi}s
%in our testbed. Those had to be used to put them in a domain name anyway.

%Below is a snippet of the file of all the hostnames and \ac{MAC} addresses:

%Below is a part of the file where we keep track of the MAC addresses and assign
%hostnames.
%This file will be available online and copied to the Raspbian lite
%image in a later step under the listed filename. For different \ac{Raspi}s
%the MAC addresses has to be changed. The MAC address of a network card can be found
%using the command \texttt{"ifconfig"} or \texttt{"ip addr"}.

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp00
b8:27:eb:7b:c3:91 rasp01
b8:27:eb:54:9c:64 rasp02
b8:27:eb:95:bd:11 rasp03
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

%Provided that each \ac{Raspi} will have knowledge of all other \ac{Raspi}s in
%the testbed, it still needs to know which name to give itself.
%We do this using a small Bash script:
%We use the \ac{Bash} script below to enable each \ac{Raspi} to find its
%hostname in \texttt{nodes.csv}.
A \ac{Bash} script can then be used to have each \ac{Raspi} assign its hostname:

% Set hostname
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/set\_hostname}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

script_path="$(dirname $(realpath $0))"
config_file=${script_path}/nodes.csv
mac=$(cat /sys/class/net/eth0/address)
old_hostname=$(hostname)
new_hostname=$(grep $mac $config_file | cut -f2 -d' ')

# Assign hostname found in nodes.csv
if [ ! -z ${new_hostname} ]; then
    echo ${new_hostname} > /etc/hostname
    hostname ${new_hostname}
    sed -i.old -e "s:${old_hostname}:${new_hostname}:g" /etc/hosts
fi
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The code lines above do the following:
\begin{itemize}
    \item Line 1: Tells system to interpret the script using \ac{Bash}
    \item Line 3-4: Gets the path to the script itself and the list of hostnames
    \item Line 5: Gets the \ac{MAC} address of the node itself
    \item Line 6: Gets the current hostname
    \item Line 7: Gets the new hostname from the hostname list
    \item Line 10-14: Assigns the new hostname
\end{itemize}

%Line 1 tells the system to interpret the script using Bash. Line 3 gets the
%path of the script itself. Line 4 gets the MAC address. This script will be
%located in all \ac{Raspi}s and thus return the MAC address of the node itself
%when executed.
%Line 4 searches line by line for this mac address.
%Line 8-11 assigns the hostname found in \texttt{"nodes.csv"}. If the MAC did not exists,
%then it will keep the default hostname of the system.

It is likely that the testbed administrator needs to either change or add
scripts to configure the \ac{Raspi}s. This would be tedious to distribute to
all \ac{Raspi}s at a later stage. We ease this task by making the \ac{Raspi}s fetch the
configuration scripts during start up before running the configuration scripts.

The below script attempts to download/update the scripts:

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/update\_rasp\_config}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

url="http://kom.aau.dk/project/TuneSCode/raspi/"
config_file="rasp_config.zip"

# Attempt to fetch new configuration files
if ! wget -q --show-progress -O /tmp/${config_file} ${url%/}/${config_file}; then
    echo "Warning: Unable to update rasp_config files"
    exit 1
fi

# Unzip and overwrite configurationn files to root's home directory
unzip -q -o /tmp/${config_file} -d /home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

\begin{itemize}
    \item Line 3-4: Is the url and zip file that should be downloaded
    \item Line 7-10: Downloads the new configuration files to \texttt{/tmp}
        and prints a warning in case of errors
    \item Line 13: Unzips the files to \texttt{/home/pi/} while ignoring existing files/directories
\end{itemize}

We have packed the files and uploaded them to a webserver. These can be
retrieved with \texttt{wget} and unzipped into the Raspbian lite image:

% Get the files
\begin{lstlisting}[]
$ wget ${URL%/*}/rasp_config.zip
$ unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For the above scripts to work in the \ac{Raspi}s, it is required that the \ac{Raspi}
\ac{MAC} address is found in \texttt{nodes.csv}. Thus,
\texttt{rasp\_config.csv} needs to be unzipped, altered, and zipped again in
the \ac{HTTP} server from which the \ac{Raspi} retrieves the files from.
The testbed administrator will therefore need to also edit the \ac{URL} in the
script \texttt{update\_rasp\_config}.
%"nodes.csv" can be changed using a text editor. For vim, this could be done
%the following way:
%\begin{lstlisting}[]
%$ vim ${ROOTDIR}/home/pi/rasp_config/nodes.csv
%\end{lstlisting}
%\FloatBarrier
%\vspace{-5mm}

Finally, to actually make the \ac{Raspi}s change hostname, we have to make
each \ac{Raspi} call the above script when it starts up. Optionally,
also run the update script before running other scripts.
Do this with root permissions by inserting the line(s) below before
\texttt{"exit 0"} using your favorite editor:
%"\$\{ROOTDIR\}/etc/rc.local" using
%your favorite text editor..
%The file
%can be opened with your favorite text editor.

\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/set_hostname
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Configuring \texttt{rc.local} makes it really cumbersome for the testbed
administrator to add new scripts since it first needs to be included
in \texttt{rasp\_config.zip} and then each \ac{Raspi} needs to have an additional
line added to \texttt{rc.local} in the root image or in all memory cards depending on
the final testbed setup.

A solution to this is to make yet another script in \texttt{rasp\_config.zip}
that calls all the other scripts in the desired order. This would look like
below:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/main}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
#!/usr/bin/env bash

bash /home/pi/rasp_config/set_hostname
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to use the above script, \texttt{rc.local} needs to look like this:
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/etc/rc.local}@>
<@\textcolor{gray}{
-----------------------------------------------------}
\Reactivatenumber @>
...
bash /home/pi/rasp_config/update_rasp_config
bash /home/pi/rasp_config/main
...
exit 0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
Notice that \texttt{set\_hostname} is now called by the \texttt{main}
script instead. The update script is still called directly. This ensures that
all configuration scripts are updated before executed. Changes to the update
script itself will first take effect at next system startup.
%Had the update script
%been called in \texttt{main} instead would mean that changes to the
%\texttt{main} script would also first have been available at next startup.

%This can be done by inserting a call to the script
%set\_hostname in rc.local before it exits.
%
%Call set\_hostname script at startup. We insert a line of code to call script in rc.local just after the initial comments (i.e. lines starting with \#).
%\begin{lstlisting}[]
%# line_number=$(egrep -n -m1 "(^[^#])|(^$)" ${ROOTDIR}/etc/rc.local | cut -d: -f1)
%# sed -i "${line_number}c bash /home/pi/rasp_config/set_hostname" ${ROOTDIR}/etc/rc.local
%\end{lstlisting}
%\FloatBarrier
%$ line="bash /home/pi/.rasp_config/set_hostname"
%# awk -v text="$line" '!/^#/ && !p {print text; p=1} 1' ${ROOTDIR}/etc/rc.local > <@ \Suppressnumber @>
%    ${ROOTDIR}/etc/rc.local <@ \Reactivatenumber @>

%WE NEED TO ADD A FEW COMMANDS TELLING HOW TO ZIP, MODIFY, AND UNZIP
%FOR THE READER TO BE ABLE TO UPLOAD A CONSTRUCT HIS/HER OWN BUNDLE.

%All other files could be updated as well.
It is not only possible to add and configure files to/within the image. One can
also install and update the packages within the image which will be illustrated
in the next section.

\subsection{Installing and updating the image}

It may be desired to pre-install some programs in the image before it is
written to all the memory cards that goes into the \ac{Raspi}s.
This can be done from any linux x86 machine using QEMU Chroot~\cite{QemuUserEmulation}
(change root).
%\url{https://wiki.debian.org/QemuUserEmulation}
%\url{https://wiki.archlinux.org/index.php/Raspberry_Pi}

Due to the \ac{ARM} processor that \ac{Raspi}s are using, it is required to
install some additional software:
%Because \ac{Raspi}s are equipped with an \ac{ARM} processor
%%it is not as
%straightforward as performing the task for two Linux systems of the same
%architecture.

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --display qemu-arm
qemu-arm (enabled):
     package = qemu-user-static
        type = magic
      offset = 0
       magic = \x7fELF\x01\x01\x01\x00\x00\x00\x00\x00\x00\x00\x00\x00\x02\x00\x28\x00
        mask = \xff\xff\xff\xff\xff\xff\xff\x00\xff\xff\xff\xff\xff\xff\xff\xff\xfe\xff\xff\xff
 interpreter = /usr/bin/qemu-arm-static
    detector =
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
% This was neccessary on arch
%# update-binfmts --importdir /var/lib/binfmts/ --import

Make sure the second command writes "enabled" as the output above. If that
is not the case, then try enabling it:

\begin{lstlisting}[]
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Provided that qemu-arm is enabled, we should now be able to change root (chroot)
into our Raspbian lite image. Change root is a method in Linux that enables
the root (/) to be changed. Thus, enabling a Linux installation within a
Linux installation.

There are a few commands to be performed before actually changing root into the
root partion of the image.
First, to get internet access from within the Raspbian lite image it is needed
to copy our \texttt{resolv.conf} into the filesystem. Line 1 change directory
into root filesystem of the image. The second line copies \texttt{resolv.conf}
from the testbed administrators \ac{PC} to the image.

% CHROOT to OS image
\begin{lstlisting}[]
$ cd $ROOTDIR
# cp /etc/resolv.conf ${ROOTDIR}/etc/resolv.conf
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, because of the \ac{ARM} architecture,
\texttt{/usr/bin/qemu-arm-static} needs to be copied into the image before continuing:

% CHROOT to OS image
\begin{lstlisting}[]
# cp /usr/bin/qemu-arm-static ${ROOTDIR}/usr/bin
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The final preparation before changing root is to populate the directories \texttt{proc},
\texttt{sys}, and \texttt{dev}:

% CHROOT to OS image
\begin{lstlisting}[]
# mount  -t proc proc proc/
# mount --bind /sys sys/
# mount --bind /dev dev/
# mount --bind /dev/pts dev/pts
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Finally, it is time to change root.
%Finally, the filesystem is ready for us to change root.
This can be done with \texttt{proot}. It may be required to install \texttt{proot}
using "\texttt{apt-get install proot}":

% CHROOT to OS image
\begin{lstlisting}[]
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Most available material online uses the more known \texttt{chroot} command as
written in the code block below instead of \texttt{proot}. This did not work
correctly in our machines, but we present the command as an alternative to
\texttt{proot} in case the testbed administrator should be in the opposite
situration where \texttt{proot} is not working.

% CHROOT to OS image
\begin{lstlisting}[]
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash    (ALTERNATIVE TO proot)
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

If everything went well, we should now be \texttt{chrooted} into the Raspbian lite
filesystem under its root. Optionally, change the prompt title to indicate that it
is a \texttt{chrooted} environment:

% Optionally, we may create a unique prompt to indicate we have changed root
\begin{lstlisting}[]
# export PS1="(chroot) $PS1"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

The Raspbian lite system image should now be possible to use almost as if it
had been booted in a \ac{Raspi}.
A major difference is that the testbed administrator's \ac{PC} is likely
significantly faster than a \ac{Raspi}.
%The only difference is that the laptop that we are using is
%significantly faster than a \ac{Raspi}.
Thus enabling updates, upgrades, and installing new software packages much
faster than in the \ac{Raspi}s.
%Thus enabling us to for example update,
%upgrade, and install software packages:

% Update system
\begin{lstlisting}[]
(chroot) # apt-get update
(chroot) # apt-get upgrade
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

We also install a few packages that we find useful:
%Lets install some useful applications:
% Install packages
\begin{lstlisting}[]
(chroot) # apt-get install vim git screen
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

When writing the image to a memory card, all these change that has been made
to the image so far will exists in all \ac{Raspi}s.
%All the changes that are made here will exists in all \ac{Raspi}s when the
%image is written to a memory card.
